### 2015-05-10 papers/histogram/thesis/thesis.tex

Throughout the discussion and development of histogram methods, we
have thus far neglected a few important subjects:
\begin{enumerate*}
\item initialization end conditions i.e. the means by which we decide
  when an algorithm to construct weights is finished,
\item the minimum temperature $T_{\t{min}}$ which we have sometimes
  mentioned,
\item the energy range over which we are interested in, and
\item what to do with the weight array outside of the energy range of
  interest, or at in which it is not defined.
\end{enumerate*}
The last three of these subjects turn out to be related, as the
minimum temperature $T_{\t{min}}$ chosen for any given simulation
determines the minimum energy $E_-$ of interest, as well as the
weights $w\p{E}$ at energies $E<E_-$.

@@

While the Wang-Landau method comes equipped with an end condition of
its own, most histogram methods require some external means of
deciding when they are finished initializing. Wang-Landau's
``natural'' end condition does not advantage it over other methods,
however, as the WL end condition makes use of a free parameter, which
any end condition will require. In order to compare all methods
fairly, we need a method-independent end condition. The most
straightforward of such end conditions is to simply to stop
initializing after some fixed number of Monte Carlo iterations, and
appropriately finish up any other calculations in the algorithm before
terminating it. We thus initialize systems of $N$ spheres for
$10^5N^2$ iterations; the choice of $10^5N^2$ we will address in
Section \ref{sec:results}. \fixme{address in results section} In
general, however, one might want an end condition which is independent
of system size. We came up with a four such end conditions as
examples, though we do not use these end conditions in this paper.

Let $E_+$ be the energy at which the density of states $D\p{E}$ and
entropy $S\p{E}$ are maximal, and let $E_-$ be the minimum energy of
interest. We define the pessimistic sample count $s_p\p{E_0}$ as the
number of times which a system has traveled from $E_+$ to $E_0$, in
any number of moves. This count represents the most pessimistic
estimate of the number of independent, uncorrelated observations of
the energy $E_0$, as going to the state of maximal entropy between
samples wipes all correlations between them. We similarly define an
optimistic sample count $s_o\p{E_0}$ as the number of times which the
system has transitioned into $E_0$ from any energy $E>E_0$. This count
represents the most optimistic estimate of the number of independent
observations of the energy $E_0$, as the system had merely to go to
any energy $E>E_0$, where the density of states may be only marginally
larger, between each sample.

In general, any initialization routine should sample the entire range
$R=\sp{E_-,E_+}$ of interest for simulation before terminating.
Furthermore, samples at the end of initialization should be numerous
enough to be statistically significant. Given that any samples of
$E_-$ necessarily require first sampling all other energies in $R$ on
the way down from $E_+$, this energy ($E_-$) will generally have the
fewest number of samples. As an end condition, therefore, one can
simply enforce some minimum number of optimistic or pessimistic
samples at $E_-$, i.e. $s_o\p{E_-}$ or $s_p\p{E_-}$. A more
sophisticated end condition would be to choose a minimum temperature
of interest $T_{\t{min}}$ and initialize until the mean fractional
error $\bk{\delta_s}_{E,T_{\t{tmin}}}$ in either of the sample counts
$s_o\p{E}$ or $s_p\p{E}$ reaches some minimum value. These mean errors
are
\begin{align}
  \bk{\delta_s}_{E,T}=\bk{s^{-1/2}}_{E,T}=\sum_Es\p{E}^{-1/2}P\p{E,T}
  =\f{\sum_Es\p{E}^{-1/2}D\p{E}e^{-E/kT}}{\sum_{E'}D\p{E'}e^{-E'/kT}},
  \label{eq:fractional_sample_error}
\end{align}
where we use the fact that the fractional error $\delta_G$ in any
histogram $G$ collected via random Monte Carlo sampling is $G^{-1/2}$.

@@

Using histogram methods generally requires knowing the energy range of
interest for the system at hand. The Wang-Landau method, for example,
periodically checks a flatness condition $C\sp{H}$ on the energy
histogram $H\p{E}$. In order for such a condition to be well-defined,
it needs have a specified energy range over which to check the
flatness of $H$. While some systems, such as the Ising model, have
well-defined and easily computable energy range, the same is not true
of most systems. In the case of the square-well fluid, not only is the
minimum energy unknown, some energies have such an incredibly low
state density $D\p{E}$ that one cannot reasonably expect to ever
observe them via Monte Carlo, biased or otherwise. We therefore need
to identify an energy range of interest, specified by a maximum and
minimum energies of interest.

Identifying a maximum energy of interest is straightforward: this is
simply the energy $E_+$ at which the density of states $D\p{E}$ is
maximal. Though high energies become more important for determining
system properties $\bk{X}_T$ at higher temperatures, the relative
importance of energies $E>E_+$ to the energy $E_+$ approaches a
maximum as $T\to\infty$. There is therefore a maximum weight $w\p{E}$
which is appropriate to assign energies $E>E_+$ relative to the weight
$w\p{E_+}$, which is namely the ratio of these weights as
$T\to\infty$. As observed in (\ref{eq:inf_temp_weights}), the weight
array becomes flat as $T\to\infty$, which means that the ratio of
$w\p{E>E_+}$ to $w\p{E}$ appropriate for any given temperature $T$
maximally approaches unity. We can therefore always set
$w\p{E>E_+}=w\p{E_+}$, which will ensure that energies $E>E_+$ are not
sampled any more than is necessary for any positive temperature. The
maximum of $D\p{E}$ can be found by simulating without weights for a
short while, after which the energy histogram $H\p{E}$ should be
maximal at $E_+$.

The minimum important energy $E_-$ for a system is less obvious than
$E_+$. Just as finding $E_+$ naturally required using a ``maximal''
temperature $T\to\infty$, determining $E_-$ requires a choice of some
minimum temperature $T_{\t{min}}$. The minimum important energy $E_-$
is then the energy at which the partition function terms
$D\p{E}e^{-E/kT_{\t{min}}}$ are maximal. Lower energy terms grow
increasingly less important with decreasing energy, and are
sufficiently sampled with canonical weights
$w\p{E<E_-}=e^{-E/kT_{\t{min}}}$ to determine any property
$\bk{X}_{T\ge T_{\t{min}}}$. To maximize $D\p{E}e^{-E/kT_{\t{min}}}$,
we need
\begin{align}
  \f{d}{dE}\p{De^{-E/kT_{\t{min}}}}
  =\f{dD}{dE}e^{-E/kT_{\t{min}}}+De^{-E/kT_{\t{min}}}\p{-\f{1}{kT_{\t{min}}}}=0,
  \label{eq:min_E_setup}
\end{align}
which means
\begin{align}
  \f1{D}\f{dD}{dE}=\f{d\ln D}{dE}=\f{1}{kT_{\t{min}}}.
  \label{eq:min_E_condition}
\end{align}
Though we do not always know the density of states $D\p{E}$, in
practice we compute the best available estimate $D\p{E}$ every time we
need to know $E_-$. If we do not find any energy satisfying
(\ref{eq:min_E_condition}), we use the minimum energy the simulation
has ever observed in place of $E_-$ to define our energy range of
interest.

Once we identify an energy $E_-$ satisfying (\ref{eq:min_E_condition})
in an initialization routine for any histogram method, we set weights
$w\p{E<E_-}=e^{-\p{E-E_-}/kT_{\t{min}}}$ with a temperature
$kT_{\t{min}}/\epsilon=0.2$. These weights prevent the initialization
routine from wasting time on energies at which we will override the
weight array after initialization anyways. We chose this temperature
because it is comfortably below the temperature at which the
square-well fluid condenses from a gas into a liquid (see Section
\ref{sec:properties}). \fixme{what about pressure?}

We made a special case for Wang-Landau treatment of the energy range
of interest in, as the design of this method assumes a priori
knowledge of $E_-$. In order to more fairly compare Wang-Landau to
other methods, we therefore also determined the appropriate energy
$E_-$ for each system size $N$ using other histogram methods, and
simulated via Wang-Landau with the a lower energy bound hard-coded. We
will distinguish the two sets of Wang-Landau simulations by referring
to those in which $E_-$ is hard-coded as ``vanilla Wang-Landau.''

@@

Having developed and implemented all of the histogram methods
discussed in Section \ref{sec:methods}, we will now look at the
results from simulating with these methods. Due to time constraints,
we simulated only one system, the square-well fluid with well width
$\lambda=1.3$ and filling fraction $\eta=0.3$, for sizes (i.e. number
of spheres) $N\in\sp{5,25}$. We initialized histogram methods for
$10^5N^2$ iterations ($10^5N^3$ moves), and simulated for just as long
afterwards. To check for consistency in simulation results, we
simulated each combination of histogram method and system size for 30
different random number generator seeds.

@@

Our analysis will involve comparing errors in system properties
computed from simulation results. As we do not know the true values of
these system properties, we will compare standard simulation results
against those from ``gold standard'' simulations, which run for
significantly longer than the standard $10^5N^2$ iterations. For our
gold standard simulations, we use the TMMC method and initialize until
it collects $10^4$ optimistic samples of the minimum important energy
(i.e. until $s_o\p{E_-}\ge 10^4$). \fixme{why TMMC?} To increase
confidence in low-temperature gold standard results, we initialized
them with $kT_{\t{min}}/\epsilon=0.1$. After initialization, the gold
standard runs simulate for $10^{10}$ iterations ($10^{10}N$
moves). While the entire initialization and simulation process takes
around two hours for the largest regular simulations, the largest
golden simulations initialize for $\sim11$ hours, and simulate for a
week. To control for iteration number in the quality of results, we
compare against ``converged flat'' simulations. Converged flat
simulations read in the transition matrix from golden simulations
(which accumulates statistics in both during and after initialization)
to construct a weight array, but simulate for the same number of
iterations as the standard simulations.

@@

Most of our analysis will use quantities computed from two arrays
given by simulations: the energy histogram $H\p{E}$ and the weights
$w\p{E}$. To get a sense of what these arrays look like, Figure
\ref{fig:arrays} shows the energy histogram and weights from an $N=25$
simulation, along with the density of states computed via $\tilde
D=H/w$ and normalized to have a maximum value of 1. The vertical
reference line in these plots denotes the minimum important energy
$E_-$. For reference, we also provide the histogram produced by an
infinite temperature (no weights) simulation.

The most immediate observation in these plots is the failure of
Wang-Landau to produce meaningful results without a predefined energy
range. This failure occurs because the Wang-Landau algorithm does not
have a mechanism for properly setting and adjusting weights on newly
observed energies. After several reset cycles of the Wang-Landau
algorithm, the weight-modifying factor $f_k$ gets exceedingly small,
causing Wang-Landau to get stuck at newly encountered low energies
trying to fix their weights with a factor $f_k$ whose magnitude is
appropriate for fine tuning, rather than heavy adjustment. If
Wang-Landau knows the energy range of interest to begin with, it will
not decrease $f_k$ until it has observed all important energies, and
adjusted their weights to some reasonable value. Due to the drastic
failure of Wang-Landau, will omit its results, which are nonsensical,
from many figures.

The energy histogram in Figure \ref{fig:arrays} shows that TMMC and
OE-TMMC spend considerably less time at important low energies than
the simple flat and vanilla Wang-Landau methods. At face value, this
behavior of TMMC and OE-TMMC is undesirable.

High energy histogram counts at low energies mean that a simulation
has collected many statistics on these energies, but the energy
histogram contains no information about how correlated different
energy observations are. To determine how effective different
simulations are at collecting uncorrelated statistics, we can look at
the pessimistic sample rate $r_p\p{E}$, which is the mean number of
simulation iterations required to sample a given energy $E$. This rate
is given in terms of the number of iterations $I$ in a simulation
(where we always have $I=10^5N^2$) and the pessimistic sample rate
$s_p\p{E}$ by $r_p\p{E}=I/s_p\p{E}$. Figure \ref{fig:sample_rates}
shows pessimistic sample rates for all methods except Wang-Landau
averaged over 30 random number generator seeds for $N=25$
simulations. Lower pessimistic sample rates mean that a simulation
takes less time to collect statistics which are guaranteed to be
uncorrelated. By this metric, TMMC and OE-TMMC again perform worse
than simple flat and vanilla Wang-Landau.\fixme{comment on OETMMC
  doing worse than TMMC}

@@

Histograms and sampling rates let us speculate about the quality of a
simulations, but the most important task for any simulation is to
yield accurate physical results. We therefore compute, via the formula
in (\ref{eq:XT}), the temperature dependence of the square-well
fluid's potential energy
\begin{align}
  U=\bk{E}_T,
  \label{eq:U}
\end{align}
the heat capacity
\begin{align}
  C_V\p{T}=\f{\partial U}{\partial T}
  =\bk{\p{\f{E}{kT}}^2}_T-\bk{\f{E}{kT}}_T^2,
  \label{eq:CV}
\end{align}
and the configurational entropy\fixme{how is this different from
  regular entropy?}
\begin{align}
  S_{\t{config}}\p{T}=k\bk{\f{E}{kT}+\ln Z}_T
  -k\bk{\f{E}{kT}+\ln Z}_{T\to\infty}.
  \label{eq:S}
\end{align}
All of these properties are directly computable from the density of
states $D\p{E}$, and are shown in Figures
\ref{fig:internal_energy}--\ref{fig:config_entropy} for $N=25$
simulations, averaged over 30 random number generator seeds. When
looking at these figures, one should remember that we are only
confident in results down to $kT/\epsilon=0.2$ for regular
simulations, and $kT/\epsilon=0.1$ for golden ones. With the exception
of Wang-Landau, which has a heavy low-energy bias, all simulations
agree fairly well on these properties of the square-well fluid.

Though Figures \ref{fig:internal_energy}--\ref{fig:config_entropy} are
interesting for studying the square-well fluid, they do not help us
resolve the efficacy of different histogram methods. These figures do,
however, show that the square-well fluid has a phase transition around
$kT/\epsilon\approx0.5$, which justifies our choice of
$kT_{\t{min}}/\epsilon=0.2$ and ensures that the histogram methods are
indeed performing the nontrivial task of exploring the liquid-gas
phase boundary of the square-well fluid. For the purposes of this
paper, we will neglect to demonstrate that this phase transition is
indeed between liquid and gas phases, and take this fact as given.

@@

Having a sense of how $U$, $C_V$ and $S_{\t{config}}$ behave with
respect to temperature, we can now look at the how the errors $\Delta
U$, $\Delta C_V$ and $\Delta S_{\t{config}}$ relative to the TMMC
golden calculations for each method (except Wang-Landau). Figures
\ref{fig:internal_energy_err}--\ref{fig:config_entropy_err} show these
errors. Much like the arrays in Figure \ref{fig:arrays}, we provide
these figures only for a sense of how error curves behave. Errors (at
$kT/\epsilon\ge0.2$) typically peak near the phase transition at
$kT/\epsilon\approx0.5$, and drop back down.

Figures \ref{fig:u_scaling}--\ref{fig:s_scaling} provide a more
comprehensive picture of errors from each method. These figures show
the average maximum error in $U$, $C_V$, and $S$ for different system
sizes. Though errors jump around considerably in these figures, the
fact that they are roughly level between $N=15$ to $25$ justifies the
choice $N^2$ scaling for our number of initialization iterations.
Future studies of larger systems might use this scaling for initial
estimates of initialization times.

@@

It is more meaningful to compare errors against errors, rather than
system size. \fixme{wording} Figures
\ref{fig:u_comps}--\ref{fig:s_comps} show the errors of each method
plotted against errors of the converged flat histogram weight
simulations. The fact that some points fall below the diagonal
reference line is strange, and indicates that some regular simulations
are outperforming the converged flat ones. We do not understand why
such events occur, as converged flat simulations should be using
high-quality golden data to construct weight arrays, and should
therefore outperform all standard simulations. These events are thus
worthwhile looking into in more detail.

Figures \ref{fig:u_scaling}--\ref{fig:s_comps} make clear, again, the
inability of Wang-Landau to sample a system without a predefined
energy range. These figures also show that the performances of simple
flat method, vanilla Wang Landau, and TMMC are about the same, except
for an anomalous case with $N=21$ in which simple flat yields
simulation results with large errors. We do not know the reason for
this anomalous case, which might be worth investigating. Finally,
these error figures further confirm earlier suspicions of OE-TMMC's
underperformance relative to TMMC, despite the fact that it is
supposed to provide an improvement over the TMMC weight array.

@@

While the results in this paper are not strongly conclusive, they may
be used as guidelines for future studies of histogram methods,
especially as applied to the square-well fluid. After implementing
several histogram methods and applying them to the $\lambda=1.3$,
$\eta=0.3$ square-well fluid, we found that Wang-Landau, together with
a predefined energy range, consistently yields the smallest errors and
highest low-energy sampling rates. Without prior knowledge of the
minimum important energy of a system, however, Wang-Landau can get
stuck at low energies, and is incapable of recovering. Wang-Landau is
therefore most appropriate for studying systems in which the entire
energy range is known a priori, or in which only a given energy range
is of particular interest.

The simple flat method and TMMC methods seem to perform roughly on par
with vanilla Wang-Landau (i.e. Wang-Landau with a specified energy
range), except an anomalous case of failure worth investigating on the
part of simple flat. Our implementation of the optimized ensemble
using a transition matrix, which is meant to improve the TMMC weight
array, seemed to only make the weight array worse.

\fixme{fix when I'm awake}

>>>>

Throughout the discussion and development of histogram methods, we
have thus far neglected expanding on a few important subjects:
\begin{enumerate*}[label=\roman*)]
\item initialization end conditions i.e. the means by which we decide
  when an algorithm to construct weights is finished,
\item the minimum temperature $T_{\t{min}}$ which we have sometimes
  mentioned,
\item the energy range over which we are interested in, and
\item what to do with the weight array outside of the energy range of
  interest, or at energies where it is not defined.
\end{enumerate*}
The last three of these subjects turn out to be related, as the
minimum temperature $T_{\t{min}}$ chosen for any given simulation
determines the minimum energy $E_-$ of interest, as well as the
weights $w\p{E}$ at energies $E<E_-$.

@@

While the Wang-Landau method comes equipped with an end condition of
its own, most histogram methods require some external means of
deciding when they are finished initializing. Wang-Landau's
``natural'' end condition does not advantage it over other methods,
however, as the WL end condition makes use of a free parameter (the
cutoff $c$), which any end condition will require. In order to compare
all methods fairly, we need a method-independent end condition to
initialization. The most straightforward of such end conditions is to
simply to stop initializing after some fixed number of Monte Carlo
iterations, and finish up any other calculations in the algorithm if
appropriate. We thus initialize systems of $N$ spheres for $10^5N^2$
iterations; the choice of $10^5N^2$ we address in Section
\ref{sec:results}. In general, however, one might want an end
condition which is independent of system size. We came up with a four
such end conditions as examples, though we do not use them in
simulations analyzed in this paper.

Our end conditions require knowledge of the minimum and maximum
energies of interest, which we denote respectively by $E_-$ and $E_+$.
We define the pessimistic sample count $s_p\p{E_0}$ as the number of
times which a system has traveled from $E_+$ to $E_0$. The trip from
$E_+$ to $E_0$ may be made in any number of moves, but cannot include
intermediate visits to $E_-$ or $E_0$. The count $s_p\p{E_0}$ provides
the most pessimistic estimate for the number of independent,
uncorrelated observations of the energy $E_0$, assuming that a visit
to $E_+$ wipes all correlations between samples of any other energy;
this assumption will be justified by our definition of $E_+$ as the
energy at which the density of states $D\p{E}$, and by extension the
entropy $S\p{E}$, is maximal. We similarly define an optimistic sample
count $s_o\p{E_0}$ as the number of times which the system has
transitioned into $E_0$ from any energy $E>E_0$ in a single move. This
count represents an optimistic estimate of the number of independent
observations of $E_0$, as the system had merely to go to any energy
$E>E_0$, where the density of states may be only marginally larger,
between each sample.

In general, any initialization routine should sample the entire range
$R=\sp{E_-,E_+}$ of interest for a given simulation before
terminating. Furthermore, samples at the end of initialization should
be numerous enough to be statistically significant. Given that any
samples of $E_-$ require first sampling all other energies in $R$ on
the way down from $E_+$, this energy ($E_-$) will generally have the
fewest number of samples. As an end condition, therefore, one can
simply enforce some minimum number of optimistic or pessimistic
samples at $E_-$, i.e. minimum values of $s_o\p{E_-}$ or
$s_p\p{E_-}$. A more sophisticated end condition would be to choose a
minimum temperature of interest $T_{\t{min}}$ and initialize until the
mean fractional error $\bk{\delta_s}_{E,T_{\t{tmin}}}$ in one of the
sample counts reaches some minimum value. These mean errors are
\begin{align}
  \bk{\delta_s}_{E,T}=\bk{s^{-1/2}}_{E,T}=\sum_Es\p{E}^{-1/2}P\p{E,T}
  =\f{\sum_Es\p{E}^{-1/2}\tilde D\p{E}e^{-E/kT}}{\sum_{E'}\tilde
    D\p{E'}e^{-E'/kT}},
  \label{eq:fractional_sample_error}
\end{align}
where we use the fact that the fractional error $\delta_G$ in any
histogram $G$ collected via random Monte Carlo sampling is $G^{-1/2}$.

@@

Using histogram methods generally requires specifying, whether
explicitly or implicitly, the energy range of interest for a given
simulation. The Wang-Landau method, for example, periodically checks a
flatness condition $C\sp{H}$ on the energy histogram $H\p{E}$. In
order for such a condition to be well-defined, it needs have a
specified energy range over which to evaluate $C\sp{H}$. While some
systems, such as the Ising model, have a well-defined and easily
computable energy range, the same is not true of most systems. In the
case of the square-well fluid, not only is the minimum energy unknown,
but some energies have such an incredibly low state density $D\p{E}$
that one cannot reasonably expect to ever observe them via Monte
Carlo, biased or otherwise. We therefore need to identify an energy
range of interest, bounded by a minimum $E_-$ and maximum $E_+$.

Identifying a maximum energy $E_+$ of interest is straightforward:
this is simply the energy at which the density of states $D\p{E}$ is
maximal. Though high energies become more important for determining
system properties $\bk{X}_T$ at higher temperatures, the relative
importance of energies $E>E_+$ to that of the energy $E_+$ approaches
a maximum as $T\to\infty$. There is therefore a maximum weight
$w\p{E}$ which is appropriate to assign energies $E>E_+$ relative to
the weight $w\p{E_+}$, which is given by the ratio of these weights as
$T\to\infty$. As observed in (\ref{eq:inf_temp_weights}), the weight
array becomes flat as $T\to\infty$, which means that the ratio of
$w\p{E>E_+}$ to $w\p{E}$ appropriate for any given temperature $T$
maximally approaches unity. We can therefore always set
$w\p{E>E_+}=w\p{E_+}$, which will ensure that energies $E>E_+$ are not
sampled any more than is necessary for any positive temperature. The
maximum of $D\p{E}$ can be found by simulating without weights for a
short while, after which the energy histogram $H\p{E}$ should be
maximal at $E_+$.

The minimum important energy $E_-$ for a system is less obvious than
$E_+$. Just as determining $E_+$ involved reference to a ``maximal''
temperature $T\to\infty$, determining $E_-$ requires a choice of some
minimum temperature $T_{\t{min}}$. The minimum important energy $E_-$
is then the energy at which the term $D\p{E}e^{-E/kT_{\t{min}}}$ in
the partition function is maximal. Lower energy terms in the partition
function grow increasingly less important as the distance
$\abs{E-E_-}$ grows, and are sufficiently sampled with canonical
weights $w\p{E<E_-}=w\p{E_-}e^{-E/kT_{\t{min}}}$ to determine any
property $\bk{X}_{T\ge T_{\t{min}}}$. To maximize
$D\p{E}e^{-E/kT_{\t{min}}}$, we need
\begin{align}
  \f{d}{dE}\p{De^{-E/kT_{\t{min}}}}
  =\f{dD}{dE}e^{-E/kT_{\t{min}}}+De^{-E/kT_{\t{min}}}\p{-\f{1}{kT_{\t{min}}}}=0,
  \label{eq:min_E_setup}
\end{align}
which means
\begin{align}
  \f1{D}\f{dD}{dE}=\f{d\ln D}{dE}=\f{1}{kT_{\t{min}}}.
  \label{eq:min_E_condition}
\end{align}
Though we do not always know the density of states $D\p{E}$, in
practice we compute the best available estimate of $D\p{E}$ every time
we need to know $E_-$. If we do not find any energy satisfying
(\ref{eq:min_E_condition}), we use the minimum energy the simulation
has ever observed in place of $E_-$ to define our energy range of
interest.

Once we identify an energy $E_-$ satisfying (\ref{eq:min_E_condition})
in an initialization routine for any histogram method, we set weights
$w\p{E<E_-}=e^{-\p{E-E_-}/kT_{\t{min}}}$ with a temperature
$kT_{\t{min}}=0.2\epsilon$. These weights prevent the initialization
routine from wasting time on energies at which we will override the
weight array after initialization anyways. We chose this temperature
because it is comfortably below the temperature at which the
square-well fluid condenses from a gas into a liquid (see Section
\ref{sec:properties}). \fixme{what about pressure?}

We made a special case for Wang-Landau treatment of the energy range
of interest, as the design of this method assumes knowledge of the
energy range over which to evaluate $C\sp{H}$. In order to more fairly
compare Wang-Landau to other methods, as well as having Wang-Landau
determine $E_-$ in the manner described above, we also simulated via
Wang-Landau with $E_-$ hard-coded for each system size $N$. We will
distinguish the two sets of Wang-Landau simulations by referring to
those in which $E_-$ is hard-coded as ``vanilla Wang-Landau.''

@@

Having developed and implemented all of the histogram methods
discussed in Section \ref{sec:methods}, we will now look at the
results from simulating with these methods. Due to time constraints,
we simulated only one system, the square-well fluid with well width
$\lambda=1.3$ and filling fraction $\eta=0.3$, for sizes (number of
spheres) $N\in\sp{5,25}$. We initialized histogram methods for
$10^5N^2$ iterations ($10^5N^3$ moves), and simulated for just as long
afterwards. To check for consistency in simulation results, we
simulated each combination of histogram method and system size with 30
different random number generator seeds.

@@

Our analysis will involve comparing errors in system properties
computed from simulation results. As we do not know the true values of
these system properties, we will compare standard simulation results
against those from ``gold standard'' simulations, which run for
significantly longer than the standard $10^5N^2$ iterations. For our
gold standard simulations, we use the TMMC method and initialize until
it collects $10^4$ optimistic samples of the minimum important energy
(i.e. until $s_o\p{E_-}\ge 10^4$). \fixme{why TMMC?} To increase
confidence in low-temperature gold standard results, we initialized
them with $kT_{\t{min}}=0.1\epsilon$. After initialization, the gold
standard runs simulate for $10^{10}$ iterations ($10^{10}N$
moves). While the entire initialization and simulation process takes
around two hours for the largest regular simulations, the largest
golden simulations initialize for $\sim11$ hours, and simulate for a
week. To control for iteration number in the quality of results, we
compare against ``converged flat'' simulations. Converged flat
simulations read in the transition matrix from golden simulations to
construct a weight array, and simulate for the same number of
iterations as the standard simulations.

@@

Most of our analysis will use quantities computed from two arrays: the
energy histogram $H\p{E}$ and the weights $w\p{E}$. To get a sense of
what these arrays look like, Figure \ref{fig:arrays} shows the energy
histogram and weights from $N=25$ simulations, along with the density
of states computed via $\tilde D=H/w$ and normalized to have a maximum
value of 1. The vertical reference line in these plots denotes the
minimum important energy $E_-$. For reference, we also provide the
histogram produced by an infinite temperature (no weights) simulation.

The most immediate observation in these plots is the failure of
Wang-Landau to produce meaningful results without a predefined energy
range. This failure occurs because the Wang-Landau algorithm does not
have a mechanism for properly setting and adjusting weights on newly
observed energies. After several reset cycles of the Wang-Landau
algorithm, the weight-modifying factor $f_k$ gets exceedingly small
relative to its initial value. When Wang-Landau finds new low
energies, therefore, the simulation gets stuck trying to fix the
weights at these energies with a factor $f_k$ whose magnitude is
appropriate for fine tuning, rather than heavy adjustment. If
Wang-Landau knows the energy range of interest to begin with, it will
not decrease $f_k$ until it has observed all important energies, and
adjusted their weights to some reasonable value. Due to the drastic
failure of Wang-Landau, will omit its nonsensical results from some
figures. \fixme{which figures?}

The energy histogram in Figure \ref{fig:arrays} shows that TMMC and
OE-TMMC spend considerably less time at important low energies than
the simple flat and vanilla Wang-Landau methods. At face value, this
behavior of TMMC and OE-TMMC is undesirable, as these are the most
interesting energies for low temperatures, and the most difficult
energies to sample.

High energy histogram counts at low energies mean that a simulation
has collected many statistics on these energies, but the energy
histogram contains no information about how correlated these
statistics are. To determine how effective different simulations are
at collecting uncorrelated statistics, we can look at the pessimistic
sample rate $r_p\p{E}$, which is the mean number of simulation
iterations required to sample a given energy $E$. This rate is roughly
equal to the number of iterations $I$ in a simulation and the
pessimistic sample rate $s_p\p{E}$ by $r_p\p{E}=I/s_p\p{E}$; the
greater the number of iterations $I$, the better this
approximation. Figure \ref{fig:sample_rates} shows pessimistic sample
rates averaged over 30 simulations with $N=25$. Lower pessimistic
sample rates mean that a simulation takes less time to collect
statistics which are guaranteed to be uncorrelated. By this metric,
TMMC and OE-TMMC again perform worse than simple flat and vanilla
Wang-Landau.\fixme{comment on why it is interesting that OETMMC doing
  worse than TMMC?}

@@

Histograms and sampling rates let us speculate about the quality of a
simulations, but the most important deliverable of any simulation is
accurate physical results. We therefore compute, via the formula in
(\ref{eq:XT}), the temperature dependence of the square-well fluid's
potential energy
\begin{align}
  U=\bk{E}_T,
  \label{eq:U}
\end{align}
the heat capacity
\begin{align}
  C_V\p{T}=\f{\partial U}{\partial T}
  =\bk{\p{\f{E}{kT}}^2}_T-\bk{\f{E}{kT}}_T^2,
  \label{eq:CV}
\end{align}
and the configurational entropy \fixme{is this correct?}
\begin{align}
  S_{\t{config}}\p{T}=k\bk{\f{E}{kT}+\ln Z}_T
  -k\bk{\f{E}{kT}+\ln Z}_{T\to\infty}.
  \label{eq:S}
\end{align}
All of these properties are directly computable from the density of
states $D\p{E}$, and are shown in Figures
\ref{fig:internal_energy}--\ref{fig:config_entropy} for $N=25$
simulations, averaged over 30 simulations. When looking at these
figures, one should remember that we are only confident in results
down to $kT=0.2\epsilon$ for regular simulations, and $kT=0.1\epsilon$
for golden ones. With the exception of Wang-Landau, which has a heavy
low-energy bias, all simulations agree fairly well on these properties
of the square-well fluid.

Though Figures \ref{fig:internal_energy}--\ref{fig:config_entropy} are
interesting for studying the square-well fluid, they do not help us
resolve the efficacy of different histogram methods. These figures do,
however, show that the square-well fluid has a phase transition around
$kT\approx0.5\epsilon$, which justifies our choice of
$kT_{\t{min}}=0.2\epsilon$ and ensures that the histogram methods are
indeed performing the nontrivial task of exploring the liquid-gas
phase boundary of the square-well fluid. For the purposes of this
paper, we will neglect to demonstrate that this phase transition is
indeed between liquid and gas phases, and take this fact as
given. \fixme{cite instead of stating as fact?}

@@

Having a sense of how $U$, $C_V$ and $S_{\t{config}}$ behave with
respect to temperature, we can now look at the how the errors $\Delta
U$, $\Delta C_V$ and $\Delta S_{\t{config}}$ relative to the TMMC
golden calculations, shown in Figures
\ref{fig:internal_energy_err}--\ref{fig:config_entropy_err}. Much like
the arrays in Figure \ref{fig:arrays}, we provide these figures for a
sense of how error curves behave. Errors (at $kT\ge0.2\epsilon$)
typically peak near the phase transition at $kT\approx0.5\epsilon$,
and drop back down.

Figures \ref{fig:u_scaling}--\ref{fig:s_scaling} provide a more
comprehensive picture of errors from each method. These figures show
the average maximum error in $U$, $C_V$, and $S_{\t{config}}$ for
different system sizes. Though errors jump around considerably in
these figures, the fact that they are roughly stable (logarithmically)
between $N=15$ to $25$ justifies the choice $N^2$ scaling for our
number of initialization iterations.  Future studies might use this
scaling to estimate appropriate initialization time for larger
systems.

@@

Figures \ref{fig:u_comps}--\ref{fig:s_comps} provide a different
visualization of the same information as in Figures
\ref{fig:u_scaling}--\ref{fig:s_scaling}, plotting the errors of all
histogram methods against those of converged flat histogram weight
simulations. The fact that some points fall below the diagonal
reference line is strange, and indicates that some regular simulations
are outperforming the converged flat ones. We do not understand why
such events occur, as converged flat simulations should be using
high-quality golden data to construct weight arrays, and should
therefore outperform all standard simulations. \fixme{comment on why
  some methods might do better than CFW?}

As with the arrays, the error comparison figures demonstrate the
inability of Wang-Landau to sample a system without a predefined
energy range. The simple flat method, vanilla Wang Landau, and TMMC
seem to give comparable errors, except for some anomalous cases with
$N=18$ and $N=21$ in which simple flat has large errors. We do not
know the reason for these anomalies, but they could occur if the
method terminates before finding energies near $E_-$, in which case it
would get stuck in simulation at low energies with canonical weights
for the minimum temperature $kT_{\t{min}}=0.2\epsilon$. Finally, these
error figures further confirm earlier suspicions of OE-TMMC's
underperformance relative to TMMC, despite the fact that it is
supposed to provide an improvement over the TMMC weight array.

@@

After implementing several histogram methods and applying them to the
$\lambda=1.3$, $\eta=0.3$ square-well fluid, we found that
Wang-Landau, if given a predefined energy range, consistently yields
some of the smallest errors and highest low-energy sampling rates of
all histogram methods, with comparable results from TMMC and the
simple flat method. Without prior knowledge of the minimum important
energy of a system, however, Wang-Landau can get stuck at low
energies, and is incapable of recovering. Wang-Landau is therefore
most appropriate for studying systems in which the entire energy range
is known a priori, or in which only a given energy range is of
particular interest. The simple flat method also sometimes fails to
initialize; if gone unnoticed, such failures can be costly in terms of
computation time and resources. Our implementation of the optimized
ensemble using the transition matrix resulted in lower sampling rates
and higher errors than the flat histogram TMMC weights. We suspect
that the optimized ensemble is failing to improve TMMC because its
theory neglects the discretization of energies, which may be
non-negligible for the systems we simulated.

\fixme{make the conclusion more satisfying?}
